---
title: "Phase 1 — Exploratory Data Analysis"
subtitle: "Theophylline single-dose oral PK (Theoph dataset)"
author: "Pramod BR"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    number-sections: true
    fig-width: 8
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
here::i_am("reports/01_EDA_Report.qmd")
source(here::here("R", "01_eda.R"))
library(kableExtra)
```

## Objectives

This report explores the **Theophylline** single-dose dataset to understand its
structure, verify data quality, and characterise the concentration–time behaviour
before any modelling. Theophylline is a **narrow-therapeutic-index** bronchodilator
with well-described **one-compartment, first-order oral** kinetics, which makes it
an ideal teaching dataset for population PK.

The questions we answer here:

- Is the dataset complete and internally consistent (data-quality checks)?
- What do the individual and population concentration–time profiles look like?
- How variable are subjects, and does **body weight** look like a covariate worth testing?

## Dataset preparation

`prepare_theoph()` tidies `datasets::Theoph`: it coerces `Subject` to an integer
`id`, standardises names, and derives the **absolute dose** in mg (`amt = dose × wt`,
since the recorded dose is in mg/kg).

```{r prepare}
theoph <- prepare_theoph()
glimpse(theoph)
```

```{r head-table}
theoph |>
  head(11) |>
  kbl(caption = "First subject — full concentration–time record",
      digits = 2) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

The dataset contains **`r dplyr::n_distinct(theoph$id)` subjects**,
**`r nrow(theoph)` observations**, and the following columns: subject `id`,
body weight `wt` (kg), `dose` (mg/kg), absolute dose `amt` (mg), `time` (h),
and `conc` (mg/L).

## Data-quality checks

`run_dq_checks()` runs a battery of integrity checks — subject count, samples per
subject, duplicate time points, missingness, negative concentrations, time
monotonicity, baseline values, and plausible weight/dose ranges.

```{r dq}
dq <- run_dq_checks(theoph)

dq |>
  kbl(caption = "Data-quality check results") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  column_spec(2, bold = TRUE) |>
  row_spec(which(dq$status == "FLAG"), background = "#FDEDEC")
```

All checks should report **PASS** — `Theoph` is a clean benchmark dataset (12
subjects × 11 samples, no missing or negative values, time strictly increasing
within each subject).

## Per-subject summary

`subject_summary()` collapses the data to one row per subject with demographics
and observed PK landmarks (observed `Cmax`, `Tmax`, last sampled time/conc).

```{r subject-summary}
ss <- subject_summary(theoph)

ss |>
  kbl(caption = "Per-subject demographics and observed PK landmarks",
      digits = 2,
      col.names = c("ID", "Wt (kg)", "Dose (mg/kg)", "Amt (mg)", "n",
                    "Tmax (h)", "Cmax (mg/L)", "Tlast (h)", "Clast (mg/L)")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r summary-stats}
ss |>
  summarise(
    `Weight (kg)`      = sprintf("%.1f (%.1f–%.1f)", mean(wt), min(wt), max(wt)),
    `Dose (mg/kg)`     = sprintf("%.2f (%.2f–%.2f)", mean(dose_mgkg), min(dose_mgkg), max(dose_mgkg)),
    `Observed Cmax`    = sprintf("%.1f (%.1f–%.1f)", mean(cmax_obs), min(cmax_obs), max(cmax_obs)),
    `Observed Tmax (h)`= sprintf("%.2f (%.2f–%.2f)", mean(tmax_obs), min(tmax_obs), max(tmax_obs))
  ) |>
  tidyr::pivot_longer(everything(), names_to = "Parameter", values_to = "Mean (range)") |>
  kbl(caption = "Population descriptive statistics — mean (range)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## Individual concentration–time profiles

Each subject shows the classic oral-absorption shape: a rising limb (absorption),
a peak, then a log-linear decline (elimination).

```{r individual-linear}
#| fig-cap: "Individual concentration–time profiles (linear scale), faceted by subject."
eda_plot_individual(theoph, semilog = FALSE)
```

On a **semi-log** scale the terminal decline is approximately linear, consistent
with first-order (one-compartment) elimination.

```{r individual-semilog}
#| fig-cap: "Individual profiles on a semi-log scale — a near-linear terminal slope supports one-compartment kinetics."
eda_plot_individual(theoph, semilog = TRUE)
```

## Population overview

The spaghetti plot overlays all subjects, exposing the spread of exposures and
the consistent absorption/elimination pattern.

```{r spaghetti}
#| fig-cap: "All subjects overlaid (spaghetti plot)."
eda_plot_spaghetti(theoph)
```

The mean (± SD) profile summarises the central tendency and the magnitude of
between-subject variability at each nominal time.

```{r mean-profile}
#| fig-cap: "Mean concentration–time profile with ± SD ribbon."
eda_plot_mean(theoph)
```

## Body weight as a candidate covariate

Because the dose is **weight-normalised** (mg/kg), every subject received an
absolute dose proportional to weight. We inspect the weight distribution and the
relationship between observed `Cmax` and weight to motivate the covariate testing
in Phase 5.

```{r weight-dist}
#| fig-cap: "Distribution of body weight across the 12 subjects."
eda_plot_weight(theoph)
```

```{r cmax-weight}
#| fig-cap: "Observed Cmax vs body weight, with a linear fit and subject labels."
eda_plot_cmax_weight(ss)
```

```{r cmax-weight-cor}
ct <- cor.test(ss$wt, ss$cmax_obs)
```

The Pearson correlation between weight and observed `Cmax` is
**r = `r round(ct$estimate, 2)`** (p = `r signif(ct$p.value, 2)`). This is
exploratory only — formal covariate evaluation happens within the population
model (Phase 5), where weight will be tested on clearance and volume.

## Key takeaways

- The dataset is **clean and complete**: 12 subjects, 11 samples each, no missing
  or negative values, time strictly increasing.
- Profiles are consistent with **one-compartment oral PK with first-order
  absorption** — the structural model carried into Phase 3.
- There is meaningful **between-subject variability** in exposure, motivating an
  IIV model (Phase 4).
- **Body weight** is a physiologically plausible covariate worth testing on
  clearance and volume (Phase 5).

> Next: **Phase 2 — Noncompartmental Analysis** with `PKNCA` to quantify Cmax,
> Tmax, AUC, and half-life per subject.
