---
title: "Phase 1 – Exploratory Data Analysis"
subtitle: "Theophylline Population PK Portfolio"
author: "Pramod BR"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: true
    theme: cosmo
    fig-width: 9
    fig-height: 6
    fig-dpi: 300
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
library(tidyverse)
library(patchwork)
library(viridis)
library(scales)
library(knitr)
library(kableExtra)
library(ggrepel)
library(here)

# Load helper functions
source(here("R", "00_setup.R"))
source(here("R", "01_eda.R"))

# Prepare dataset
theoph <- prepare_theoph()

# Save raw data copy
write_csv(theoph, here("data", "raw", "theoph_raw.csv"))
```

---

## Scientific Background

### Theophylline as a PK Model Drug

**Theophylline** (1,3-dimethylxanthine) is a bronchodilator used in the management of
asthma and chronic obstructive pulmonary disease (COPD). It acts as a non-selective
phosphodiesterase inhibitor, increasing intracellular cAMP and causing smooth muscle relaxation
in the airways.

From a **pharmacokinetic perspective**, theophylline is an ideal model compound because:

- It exhibits **linear (first-order) PK** at therapeutic doses
- It is well-absorbed orally with nearly complete bioavailability
- It follows **one-compartment** kinetics with reasonable approximation
- It has a **narrow therapeutic index** (10–20 mg/L), making PK/PD relationships clinically relevant
- It shows **significant between-subject variability** in clearance, driven by covariates like
  body weight, age, smoking status, and hepatic function

### Clinical Relevance

The therapeutic range for theophylline is **10–20 mg/L**. Concentrations below 10 mg/L
are sub-therapeutic; above 20 mg/L, toxicity (nausea, arrhythmias, seizures) becomes a
serious concern. This makes individualized dosing — the goal of population PK — clinically critical.

---

## Dataset Description

### Study Design

```{r dataset-overview}
glimpse(theoph)
```

The **Theoph** dataset (Boeckmann, Sheiner & Beal, 1992) is a classic single-dose oral PK
study:

- **12 healthy adult subjects** received oral theophylline
- Dose was **weight-normalised**: 4.02–5.86 mg/kg (variable between subjects)
- **11 blood samples** per subject collected over 0–25 hours post-dose
- Plasma theophylline concentrations measured (mg/L)

### Variable Dictionary

```{r variable-table}
tibble(
  Variable = c("ID", "TIME", "DV", "AMT", "WT", "DOSE", "EVID", "MDV"),
  Type     = c("Factor", "Numeric", "Numeric", "Numeric", "Numeric", "Numeric", "Integer", "Integer"),
  Units    = c("—", "hours", "mg/L", "mg", "kg", "mg/kg", "0/1", "0/1"),
  Description = c(
    "Subject identifier (1–12)",
    "Time after oral dose administration",
    "Observed plasma theophylline concentration",
    "Administered dose in absolute mg (DOSE × WT)",
    "Body weight",
    "Weight-normalised dose",
    "Event ID: 1 = dosing record, 0 = observation",
    "Missing DV flag: 1 = dose record (no concentration), 0 = observation"
  )
) |>
  kbl(caption = "Variable dictionary for the Theoph PK dataset") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  column_spec(1, bold = TRUE, color = "#2C3E50")
```

---

## Data Quality Checks

```{r dq-checks}
dq <- run_dq_checks(theoph)

tibble(
  Check   = c("Subjects", "Total observations", "Time range (h)",
               "Concentration range (mg/L)", "BLQ observations",
               "Missing concentrations", "Missing times",
               "Weight range (kg)", "Dose range (mg/kg)",
               "Total dose range (mg)"),
  Result  = c(
    dq$n_subjects,
    dq$n_obs_total,
    paste(round(dq$time_range, 1), collapse = " – "),
    paste(round(dq$conc_range, 2), collapse = " – "),
    dq$n_below_lloq,
    dq$n_missing_conc,
    dq$n_missing_time,
    paste(round(dq$weight_range, 1), collapse = " – "),
    paste(round(dq$dose_range, 2), collapse = " – "),
    paste(round(dq$dose_mg_range, 1), collapse = " – ")
  )
) |>
  kbl(caption = "Data quality summary") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  column_spec(1, bold = TRUE)
```

::: {.callout-note}
**No missing values or BLQ observations detected.** The dataset is complete and clean —
ideal for a portfolio demonstration. In a real clinical study, BLQ handling (e.g., M1/M2
method) and imputation strategies would be pre-specified in the analysis plan.
:::

---

## Subject Characteristics

```{r subject-summary-table}
ss <- subject_summary(theoph)

ss |>
  rename(
    `Subject ID`  = ID,
    `Weight (kg)` = WT,
    `Dose (mg/kg)`= DOSE,
    `Dose (mg)`   = AMT_mg,
    `Cmax (mg/L)` = Cmax,
    `Tmax (h)`    = Tmax,
    `Clast (mg/L)`= Clast,
    `Tlast (h)`   = Tlast,
    `N obs`       = n_obs
  ) |>
  kbl(caption = "Subject-level PK summary (observed data)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  row_spec(0, bold = TRUE, background = "#2C3E50", color = "white")
```

```{r subject-chars-plot, fig.height=5}
#| fig-cap: "Body weight and dose distribution across the 12 subjects"
plot_weight_dist(theoph)
```

::: {.callout-tip title="Scientific Interpretation"}
**Body weight** ranges from `r round(min(ss$WT), 1)` to `r round(max(ss$WT), 1)` kg
(CV ≈ `r round(sd(ss$WT)/mean(ss$WT)*100, 0)`%). Because dosing was weight-normalised
(mg/kg), the absolute dose (mg) varies proportionally with weight. This weight-based dosing
strategy is intended to reduce between-subject variability in exposure — but as we will see
in the covariate analysis (Phase 5), weight may still explain residual variability in
clearance and volume of distribution.
:::

---

## Concentration-Time Profiles

### Individual Profiles (Linear Scale)

```{r individual-linear, fig.height=8}
#| fig-cap: "Individual theophylline concentration-time profiles (linear scale)"
p_ind_lin <- plot_individual_linear(theoph)
p_ind_lin
save_fig(p_ind_lin, "01_individual_linear", w = 10, h = 8)
```

::: {.callout-tip title="Scientific Interpretation"}
Key observations from individual profiles:

- **Absorption phase** (0–2 h): rapid rise to Cmax visible in all subjects — consistent
  with oral immediate-release formulation
- **Peak concentrations (Cmax)**: range from
  `r round(min(ss$Cmax), 1)` to `r round(max(ss$Cmax), 1)` mg/L, indicating substantial
  between-subject variability despite weight-adjusted dosing
- **Time to peak (Tmax)**: ranges from
  `r round(min(ss$Tmax), 1)` to `r round(max(ss$Tmax), 1)` h, suggesting variability
  in absorption rate (Ka) between subjects
- **Elimination phase**: gradual, smooth decline after Cmax — consistent with first-order
  monoexponential elimination
- **Subject 1** shows notably higher exposure; **Subject 6** shows a relatively lower Cmax —
  these differences motivate the formal PopPK model
:::

### Spaghetti Plot – All Subjects Overlaid

```{r spaghetti, fig.height=5}
#| fig-cap: "Spaghetti plot showing between-subject variability in theophylline PK"
p_spag <- plot_spaghetti(theoph)
p_spag
save_fig(p_spag, "01_spaghetti", w = 9, h = 5)
```

::: {.callout-tip title="Scientific Interpretation"}
The spaghetti plot makes **between-subject variability (BSV)** immediately visible:

- Cmax varies ~2-fold across subjects
- Tmax varies from ~1 to ~3 hours
- The elimination slopes appear broadly parallel on linear scale, suggesting a **similar
  elimination rate** (ke) across subjects — the variability is more in Cmax/Tmax than in
  the terminal slope
- This pattern is consistent with a **one-compartment oral model** where Ka and CL/V vary
  between subjects — the target of our PopPK model in Phase 3
:::

### Mean ± SD Profile

```{r mean-profile, fig.height=5}
#| fig-cap: "Mean (±SD) concentration-time profile across all 12 subjects"
p_mean <- plot_mean_profile(theoph)
p_mean
save_fig(p_mean, "01_mean_profile", w = 9, h = 5)
```

::: {.callout-tip title="Scientific Interpretation"}
The mean profile reveals the **typical PK behaviour** of theophylline:

- Mean Cmax ≈ `r round(mean(ss$Cmax), 1)` mg/L occurring at ≈ `r round(mean(ss$Tmax), 1)` h
- The SD ribbon widens at Cmax, indicating this is where BSV is largest (driven by
  variability in both Ka and CL)
- The SD ribbon narrows in the terminal phase, suggesting the elimination rate is relatively
  consistent — this will be reflected in lower IIV on CL compared to Ka
- **Mean Cmax of `r round(mean(ss$Cmax), 1)` mg/L** sits within the therapeutic window
  (10–20 mg/L) — a reassuring sign that the weight-based dosing strategy is appropriate
  for the typical subject
:::

### Semi-Log Profiles (Terminal Phase Analysis)

```{r semilog, fig.height=8}
#| fig-cap: "Semi-log concentration-time profiles. Linear terminal phase confirms first-order elimination."
p_semi <- plot_individual_semilog(theoph)
p_semi
save_fig(p_semi, "01_semilog_individual", w = 10, h = 8)
```

::: {.callout-tip title="Scientific Interpretation — Why Semi-Log Matters"}
The semi-log transformation is a **fundamental PK diagnostic tool**:

- On a semi-log scale, **first-order (monoexponential) elimination appears as a straight line**
- All 12 subjects show approximately linear terminal phases — confirming **first-order
  elimination kinetics** and validating the use of a one-compartment model
- The **slope of the terminal line = −λz (elimination rate constant)**, from which
  half-life is derived: t½ = ln(2)/λz ≈ 0.693/λz
- Minor deviations from linearity (e.g., Subject 3, early time points) may reflect the
  absorption phase not being fully resolved — this is the **flip-flop phenomenon** risk
  when Ka ≈ Ke
- A **two-compartment model** would show a biphasic profile (two distinct slopes) —
  the absence of this here supports a one-compartment assumption
:::

---

## Covariate Exploration

```{r cmax-wt, fig.height=5}
#| fig-cap: "Cmax vs body weight — preliminary covariate screening"
p_cov <- plot_cmax_vs_wt(theoph)
p_cov
save_fig(p_cov, "01_cmax_vs_wt", w = 7, h = 5)
```

::: {.callout-tip title="Scientific Interpretation"}
The scatter plot of **Cmax vs body weight** provides an early signal:

- A **negative trend** is expected if heavier subjects have larger volume of distribution
  (diluting the drug more) — this is consistent with allometric scaling principles
- A **positive trend** could indicate inadequate weight-normalisation of dose
- The regression line and confidence band here give a visual (informal) signal —
  formal statistical testing of weight as a covariate on CL and V will be done in Phase 5
- Point size encodes absolute dose: heavier subjects received more drug (mg) despite the
  same mg/kg dose, helping distinguish dose effects from true weight effects
:::

---

## Summary of EDA Findings

```{r eda-summary-table}
tibble(
  Finding = c(
    "Dataset completeness",
    "Sample size",
    "Dosing strategy",
    "Cmax range",
    "Tmax range",
    "Elimination",
    "Compartment structure",
    "Between-subject variability",
    "Covariate signal"
  ),
  Observation = c(
    "No missing values, no BLQ concentrations",
    "12 subjects × 11 time points = 132 observations",
    "Weight-normalised (mg/kg) single oral dose",
    paste0(round(min(ss$Cmax),1), "–", round(max(ss$Cmax),1), " mg/L"),
    paste0(round(min(ss$Tmax),1), "–", round(max(ss$Tmax),1), " h"),
    "First-order, confirmed by linear terminal semi-log profile",
    "One-compartment model supported (no biphasic profile)",
    "~2-fold variability in Cmax; variable Tmax suggests IIV on Ka",
    "Body weight warrants formal evaluation as covariate on CL and V"
  ),
  `Modeling Implication` = c(
    "No pre-processing required; proceed directly to NCA and modeling",
    "Adequate for PopPK; limits covariate power (small N)",
    "Include weight and dose as dataset variables; test WT on CL/V",
    "Therapeutic window (10–20 mg/L) mostly achieved",
    "IIV on Ka required in the base model",
    "Use first-order absorption in ODE model",
    "Start with 1-CMT; test 2-CMT if residuals suggest misspecification",
    "Random effects on CL, V, Ka essential",
    "Phase 5 will formally test WT → CL and WT → V"
  )
) |>
  kbl(caption = "EDA findings and modeling implications") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) |>
  column_spec(1, bold = TRUE, color = "#2C3E50") |>
  column_spec(3, italic = TRUE, color = "grey30")
```

---

## Reproducibility

```{r session-info}
#| code-fold: true
sessionInfo()
```

---

## Git Commit Recommendation

```bash
git add reports/01_EDA_Report.qmd R/01_eda.R data/raw/theoph_raw.csv outputs/figures/01_*
git commit -m "feat: Phase 1 – EDA report with concentration-time profiles and covariate screening"
git push origin main
```

---

*Next: [Phase 2 – Noncompartmental Analysis](02_NCA_Report.html)*
