---
title: "Phase 2 — Noncompartmental Analysis (NCA)"
subtitle: "Per-subject exposure metrics with PKNCA"
author: "Pramod BR"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    number-sections: true
    fig-width: 8
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
here::i_am("reports/02_NCA_Report.qmd")
source(here::here("R", "02_nca.R"))
library(kableExtra)
```

## Objectives

Noncompartmental analysis (NCA) summarises drug exposure **without assuming a
structural model**. It provides model-independent benchmarks — peak concentration,
exposure (AUC), and terminal half-life — that we will later compare against the
population-model estimates (Phases 3–5).

Parameters computed per subject:

- **Cmax / Tmax** — observed peak concentration and its time.
- **AUClast** — area under the curve to the last measurable concentration
  (linear-up/log-down trapezoidal).
- **AUCinf(obs)** — AUClast plus the extrapolated tail (Clast / λz).
- **t½** — terminal half-life, from the terminal rate constant **λz**.
- **CL/F, Vz/F** — apparent clearance and volume (dose-normalised).

## Method (PKNCA)

`run_nca()` builds a `PKNCAconc` object (concentration–time, grouped by subject),
a `PKNCAdose` object (the single oral dose in mg at t = 0), defines one **0 → ∞**
interval requesting the parameters above, and runs `pk.nca()`. PKNCA selects the
terminal-phase points for λz automatically by maximising the adjusted R².

```{r run-nca}
df  <- load_pk_data()
nca <- run_nca(df)
```

## Per-subject NCA parameters

```{r param-table}
nca$wide |>
  dplyr::select(id, cmax, tmax, auclast, aucinf.obs, half.life,
                cl.obs, vz.obs, aucpext.obs) |>
  kbl(
    caption = "Noncompartmental parameters by subject",
    digits  = 2,
    col.names = c("ID", "Cmax (mg/L)", "Tmax (h)", "AUClast (mg·h/L)",
                  "AUCinf (mg·h/L)", "t½ (h)", "CL/F (L/h)", "Vz/F (L)",
                  "AUC %extrap")
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

The **AUC %extrapolated** column is a quality flag — values well under 20%
indicate AUCinf is reliable (little of the exposure comes from extrapolation).

## Summary statistics

`nca_param_summary()` reports the **geometric mean** and **geometric CV%** —
the conventional summaries for PK parameters, which are log-normally distributed.

```{r summary-table}
smry <- nca_param_summary(nca$wide)
smry |>
  kbl(caption = "NCA parameter summary (geometric mean, CV%, range)",
      digits = 2) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r key-values}
#| include: false
thalf_gm <- round(geo_mean(nca$wide$half.life), 1)
cl_gm    <- round(geo_mean(nca$wide$cl.obs), 2)
v_gm     <- round(geo_mean(nca$wide$vz.obs), 1)
```

For theophylline, the literature half-life in healthy adults is roughly **6–9 h**;
this NCA gives a geometric-mean **t½ ≈ `r thalf_gm` h**, apparent clearance
**CL/F ≈ `r cl_gm` L/h**, and apparent volume **Vz/F ≈ `r v_gm` L** — all
physiologically sensible and a useful sanity check before population modelling.

## Visualisations

```{r fig-halflife}
#| fig-cap: "Terminal half-life by subject (dashed line = geometric mean)."
nca_plot_halflife(nca$wide)
```

```{r fig-auc}
#| fig-cap: "AUClast vs AUCinf — proximity to the identity line indicates minimal extrapolation."
nca_plot_auc(nca$wide)
```

```{r fig-distributions}
#| fig-cap: "Distribution of NCA parameters across the 12 subjects."
nca_plot_distributions(nca$wide)
```

```{r fig-cl-weight}
#| fig-cap: "Apparent clearance vs body weight — an exploratory look ahead to the covariate model."
nca_plot_cl_weight(nca$wide, df)
```

```{r cl-weight-cor}
#| include: false
wt <- df |> dplyr::distinct(id, wt)
d  <- dplyr::left_join(nca$wide, wt, by = "id")
ct <- cor.test(d$wt, d$cl.obs)
```

The Pearson correlation between weight and NCA clearance is
**r = `r round(ct$estimate, 2)`** (p = `r signif(ct$p.value, 2)`) — suggestive,
and formally tested inside the population model in Phase 5.

## Key takeaways

- NCA gives **model-independent** exposure metrics that frame the modelling work.
- Geometric-mean **t½ ≈ `r thalf_gm` h**, **CL/F ≈ `r cl_gm` L/h**,
  **Vz/F ≈ `r v_gm` L** are consistent with known theophylline PK.
- Low AUC %extrapolation means AUCinf (and therefore CL/Vz) is well determined.
- Between-subject spread in CL and AUC reinforces the need for an **IIV model**
  (Phase 4) and motivates testing **weight** as a covariate (Phase 5).

> Next: **Phase 3 — Base PopPK model** (one-compartment oral, first-order
> absorption) with `nlmixr2` / `rxode2`.
