---
title: "Phase 3 — Base Population PK Model"
subtitle: "One-compartment oral, first-order absorption (nlmixr2 / rxode2)"
author: "Pramod BR"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    number-sections: true
    fig-width: 8
    fig-height: 6
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
here::i_am("reports/03_Base_Model_Report.qmd")
source(here::here("R", "03_base_model.R"))
library(kableExtra)
```

## Objectives

Having characterised the data (Phase 1) and computed model-free exposure metrics
(Phase 2), we now fit a **structural population PK model**: a one-compartment model
with first-order oral absorption. This phase establishes the model skeleton and
the typical (population-average) parameters; between-subject variability and
covariates are layered on in Phases 4–5.

## From tidy data to NONMEM events

Estimation engines treat every row as an **event** — either a dose or an
observation. `build_nonmem_data()` inserts one dosing row per subject at
time 0 and keeps the concentration rows as observations.

```{r build-data}
dat <- build_nonmem_data()
dat |>
  dplyr::filter(ID == 1) |>
  kbl(caption = "Subject 1 as an event dataset (dose row first, then observations)",
      digits = 2) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

Column meanings: **EVID** = 1 marks the dose (with **AMT** in mg) and 0 marks an
observation (with **DV**, the concentration); **MDV** = 1 means "no measured DV on
this row" (true for dose rows); **CMT** routes the dose to the depot (1) and the
observation to the central compartment (2).

## The structural model

The model is a function with an `ini` block (starting values) and a `model` block
(the structure). Fixed effects are estimated on the **log scale**, so the
back-transformed CL, V, and Ka are always positive. Initial values are seeded from
the Phase 2 NCA, which puts the optimiser close to the solution from the start.

The fitted model's structure is shown as equations in the next section.

The two `d/dt(...)` lines are the compartment mass-balance equations; `cp = center/v`
is the predicted concentration; and `cp ~ add(add.sd) + prop(prop.sd)` specifies a
**combined additive + proportional** residual error model. The base model carries
between-subject variability on **CL only** (`eta.cl`) — variability on V and Ka is
added in Phase 4.

## Fitting the model (FOCEi)

`fit_base_model()` estimates the parameters by **FOCEi** (First-Order Conditional
Estimation with interaction), the workhorse algorithm of population PK.

```{r fit-model}
#| cache: true
#| output: false
fit <- fit_base_model(dat)
```

```{r}
#| echo: false
fit
```

## Parameter estimates

```{r param-table}
base_param_table(fit) |>
  kbl(caption = "Base model fixed-effect estimates (back-transformed)",
      digits = 3) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r compare-nca}
#| include: false
# Typical CL/V/Ka from the fixed effects (estimated on the log scale).
th <- tryCatch(fit$theta, error = function(e) nlmixr2::fixef(fit))
gp <- function(nm) if (nm %in% names(th)) as.numeric(th[[nm]]) else NA_real_
cl_hat <- exp(gp("tcl")); v_hat <- exp(gp("tv")); ka_hat <- exp(gp("tka"))
```

### Sanity check vs NCA

The population model and the model-free NCA should tell a consistent story:

| Parameter | NCA (geo. mean, Phase 2) | Base model (typical) |
|-----------|--------------------------|----------------------|
| CL (L/h)  | ~2.74                    | `r round(cl_hat, 2)` |
| V (L)     | ~31.6                    | `r round(v_hat, 1)`  |
| Ka (1/h)  | ~1.5 (from Tmax)         | `r round(ka_hat, 2)` |

Close agreement is the key validation that the structural model and dataset are
correct — the two independent methods converge on the same physiology.

## Quick predicted-vs-observed check

Full diagnostics (GOF, VPC, CWRES) come in Phase 6; here we just confirm the fit is
sensible — points should scatter around the line of identity.

```{r obs-pred}
#| fig-cap: "Population (PRED) and individual (IPRED) predictions vs observations."
fitdf <- as.data.frame(fit)

p <- ggplot2::ggplot(fitdf, ggplot2::aes(PRED, DV)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_point(alpha = 0.6, colour = "#2C3E50") +
  ggplot2::labs(x = "Population prediction (mg/L)", y = "Observed (mg/L)",
                title = "Observed vs population prediction (base model)")
p
save_fig(p, "03_base_obs_vs_pred")
```

## Key takeaways

- The data were re-expressed as a NONMEM **event** dataset (dose + observation rows).
- A **one-compartment oral** model with first-order absorption was fitted by **FOCEi**.
- Typical estimates (**CL ≈ `r round(cl_hat, 2)` L/h**, **V ≈ `r round(v_hat, 1)` L**,
  **Ka ≈ `r round(ka_hat, 2)` /h**) agree with the NCA — the structural model holds.
- The base model includes IIV on **CL** only; the next phase adds variability on
  **V and Ka**.

> Next: **Phase 4 — IIV model**, extending between-subject variability to V and Ka
> and examining the random-effect (η) distributions.
```
