---
title: "Phase 4 — Inter-Individual Variability (IIV)"
subtitle: "Adding random effects on CL, V, and Ka"
author: "Pramod BR"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    number-sections: true
    fig-width: 8
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
here::i_am("reports/04_IIV_Model_Report.qmd")
source(here::here("R", "04_iiv_model.R"))
library(kableExtra)
```

## Objectives

The base model (Phase 3) allowed only **clearance** to vary between subjects. Here
we add **between-subject random effects** on **V** and **Ka** as well, so each
subject has their own clearance, volume, *and* absorption rate. We then test
whether the extra variability is justified — better fit, well-estimated, and
plausibly distributed.

## What a random effect is

For each parameter we write, e.g., `cl = exp(tcl + eta.cl)`, where `eta.cl` is a
subject-specific deviation drawn from a normal distribution `N(0, omega^2)`. The
population shares one typical curve; each subject's curve is a shifted version of
it. The spread of those shifts — the IIV — is summarised as a **CV%**.

The fitted model's structure is shown as equations in the next section.

## Fit the IIV model

```{r fit-iiv}
#| cache: true
dat      <- build_nonmem_data()
base_fit <- readRDS(fs::path(PATH_MODELS, "base_fit.rds"))
iiv_fit  <- fit_iiv_model(dat)
iiv_fit
```

## Parameter estimates

```{r iiv-params}
base_param_table(iiv_fit) |>
  kbl(caption = "IIV model estimates (fixed effects, IIV CV%, and shrinkage)",
      digits = 3) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

Read the **BSV(CV%)** column for the size of each random effect and the
**Shrink(SD)%** column for how well the data support it. As a rule of thumb,
shrinkage under ~30% means the individual estimates are genuinely informed by the
data rather than collapsing to the population mean.

## Is the added variability justified?

We compare the base and IIV models by objective function value (**OFV**), and by
**AIC/BIC**, which penalise extra parameters. By the likelihood-ratio test, adding
two random effects (V and Ka) is significant at p < 0.05 if the OFV drops by more
than **5.99** (χ², 2 df).

```{r compare}
cmp <- compare_models(
  model_metrics(base_fit, "Base (IIV on CL)"),
  model_metrics(iiv_fit,  "IIV on CL, V, Ka")
)
cmp |>
  kbl(caption = "Base vs IIV model comparison", digits = 2) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r compare-text}
#| include: false
dofv <- cmp$dOFV[2]
daic <- cmp$dAIC[2]
```

The IIV model lowers the OFV by **`r abs(round(dofv,2))`** points relative to base
(ΔAIC = `r round(daic,2)`). A drop well beyond the 5.99 threshold means the
between-subject variability on V and Ka is **statistically warranted** — the model
genuinely fits better, not just by added flexibility.

## Random-effect distributions

The empirical-Bayes η estimates should be centred on 0 and roughly symmetric — the
model's normality assumption.

```{r eta-plot}
#| fig-cap: "Eta distributions for CL, V, and Ka (centred on zero = assumption met)."
iiv_plot_eta(iiv_fit)
```

## Key takeaways

- Random effects on **V and Ka** were added to the base (CL-only) model.
- The OFV fell by **`r abs(round(dofv,2))`** (≫ the 5.99 cut-off for 2 df), so the
  extra IIV is **justified**.
- Shrinkage and the η distributions indicate the random effects are
  data-supported, not artefacts.
- This full-IIV model is now the reference onto which we test a **covariate**
  (body weight) next.

> Next: **Phase 5 — Covariate model**, testing whether body weight explains part of
> the between-subject variability in CL and V.
```
