---
title: "Phase 5 — Covariate Model (Body Weight)"
subtitle: "Allometric weight effect on clearance and volume"
author: "Pramod BR"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    number-sections: true
    fig-width: 8
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
here::i_am("reports/05_Covariate_Model_Report.qmd")
source(here::here("R", "05_covariate_model.R"))
library(kableExtra)
```

## Objectives

The IIV model (Phase 4) left between-subject variability as purely *random*. Here
we test whether **body weight** explains part of it, by adding an **allometric**
weight effect on clearance and volume. The questions: does it improve the fit, and
does the random IIV on CL/V shrink?

## How weight enters the model

We use the pharmacometric-standard **allometric** form, normalised to a 70 kg
reference, with the exponents estimated from the data (started at the theoretical
0.75 for CL and 1.0 for V):

$$ CL = CL_{typ}\,(WT/70)^{\theta_{CL}}, \qquad V = V_{typ}\,(WT/70)^{\theta_{V}} $$

```{r show-model}
#| code-fold: false
print(cov_model)
```

## Fit the covariate model

```{r fit-cov}
#| cache: true
dat     <- build_nonmem_data()
iiv_fit <- load_iiv_fit()
cov_fit <- fit_cov_model(dat)
cov_fit
```

## Parameter estimates

The two new parameters are **dWTCL** and **dWTV** — the weight exponents. Their
95% confidence intervals tell us whether the effect is distinguishable from zero
(no effect).

```{r cov-params}
base_param_table(cov_fit) |>
  kbl(caption = "Covariate model estimates (note dWTCL, dWTV exponents)",
      digits = 3) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## Is the weight effect justified?

```{r compare}
cmp <- compare_models(
  model_metrics(iiv_fit, "IIV (no covariate)"),
  model_metrics(cov_fit, "Weight on CL & V")
)
cmp |>
  kbl(caption = "IIV vs covariate model", digits = 2) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r verdict}
#| include: false
dofv <- cmp$dOFV[2]
justified <- !is.na(dofv) && dofv < -5.99
```

Adding two parameters is significant (p < 0.05) if the OFV drops by more than
**5.99**. Here ΔOFV = **`r round(dofv, 2)`**, so on the likelihood-ratio test the
weight effect is **`r if (justified) "statistically supported" else "not clearly supported"`**
in this dataset.

## Did the random IIV shrink?

If weight genuinely drives part of the between-subject variability, the BSV(CV%) on
CL and V should be smaller in the covariate model than in the IIV model.

```{r cv-change}
dplyr::full_join(
  iiv_cv(iiv_fit) |> dplyr::rename(`IIV model` = `BSV CV%`),
  iiv_cv(cov_fit) |> dplyr::rename(`Covariate model` = `BSV CV%`),
  by = "Parameter"
) |>
  kbl(caption = "Between-subject CV% before vs after adding weight",
      digits = 1) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## Residual random effects vs weight

After including weight, the random effects on CL and V should show little residual
trend with weight — a flat line means the covariate has absorbed the size effect.

```{r eta-wt}
#| fig-cap: "Random effects on CL and V vs body weight in the covariate model."
cov_plot_eta_wt(cov_fit, dat)
```

## Key takeaways

- Body weight was added **allometrically** on CL and V, with estimated exponents.
- On the likelihood-ratio test the effect was
  **`r if (justified) "supported" else "not clearly supported"`**
  (ΔOFV = `r round(dofv, 2)`).
- We reported the change in random CV% and the residual η–weight trend honestly —
  in a small, weight-dosed dataset like Theoph a strong weight signal is not
  guaranteed, and recognising that is part of sound covariate analysis.

> Next: **Phase 6 — Model diagnostics**: goodness-of-fit plots, CWRES vs
> TIME/PRED, and a visual predictive check (VPC) on the selected model.
```
