---
title: "Phase 6 — Model Diagnostics"
subtitle: "Goodness-of-fit, CWRES, and Visual Predictive Check"
author: "Pramod BR"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    number-sections: true
    fig-width: 9
    fig-height: 8
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
here::i_am("reports/06_Model_Diagnostics.qmd")
source(here::here("R", "06_diagnostics.R"))
library(kableExtra)
```

## Objectives

A model that converges is not necessarily a model that *fits*. This phase
stress-tests the selected model with the standard pharmacometric diagnostics:
goodness-of-fit (GOF) scatter plots, conditional weighted residuals (CWRES), and a
Visual Predictive Check (VPC).

```{r load-fit}
fit <- load_final_fit()
```

## Goodness-of-fit panel

Four plots, read together:

- **Observed vs Population prediction (PRED)** and **vs Individual prediction
  (IPRED)** — points should fall along the dashed line of identity. IPRED hugs the
  line more tightly because it uses each subject's own random effects; that gap
  between PRED and IPRED is exactly what the IIV captures.
- **CWRES vs TIME** and **CWRES vs PRED** — conditional weighted residuals should
  scatter randomly in a flat horizontal band, mostly within **±2**, with no trend
  or funnel shape. A slope or curve over time would signal a structural
  misspecification; fanning would signal a wrong residual-error model.

```{r gof}
#| fig-cap: "Four-panel goodness-of-fit diagnostics for the selected model."
gof_panel(fit)
```

## Visual Predictive Check

The VPC simulates many virtual trials from the fitted model and overlays the
**observed** data on the **simulated** prediction intervals. If the observed median
and spread fall within the simulated bands across time, the model reproduces both
the typical profile and the between-subject variability — the key qualification for
using it to simulate (Phase 7).

```{r vpc}
#| fig-cap: "Visual Predictive Check: observed data vs model-simulated prediction intervals."
vpc <- make_vpc(fit, n = 500)
if (!is.null(vpc)) {
  vpc
} else {
  cat("VPC requires the 'vpc' package. Install with renv::install('vpc'), then re-render.")
}
```

## Key takeaways

- **GOF**: observed-vs-predicted should track the identity line, tighter for IPRED.
- **CWRES**: a flat, ±2 band with no time/PRED trend indicates the structural and
  error models are adequate.
- **VPC**: observed percentiles inside the simulated intervals confirm the model
  captures central tendency *and* variability.
- Passing these checks is what licenses the model for the **clinical simulations**
  in the next phase.

> Next: **Phase 7 — Simulation**: use the qualified model to simulate PK profiles
> for a typical patient and weight-based subgroups.
```
