---
title: "Phase 7 — Clinical Simulation"
subtitle: "Typical patient, weight subgroups, and population variability (rxode2)"
author: "Pramod BR"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    number-sections: true
    fig-width: 8
    fig-height: 6
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
here::i_am("reports/07_Simulation_Report.qmd")
source(here::here("R", "07_simulation.R"))
library(kableExtra)
```

## Objectives

With a model that passed diagnostics (Phase 6), we can now **simulate** — predict
concentration-time profiles for scenarios we never measured. This is where a
population PK model earns its keep clinically. We simulate three things with
`rxode2`: a typical patient, an illustrative weight effect, and the full
population spread.

```{r pars}
pars <- get_typical_params()
dose <- rep_dose()
```

The **selected model** is the **IIV model** from Phase 4 — Phase 5 found body weight
was not statistically supported, so by parsimony the IIV model is the final model.
Its typical parameters are CL ≈ `r round(pars$cl, 2)` L/h, V ≈ `r round(pars$v, 1)` L,
Ka ≈ `r round(pars$ka, 2)` /h. Simulations use a representative
**`r dose` mg** dose (the median actual dose in the dataset).

## Typical patient

Setting the random effects to zero gives the single population-average curve for a
70 kg subject.

```{r typical}
#| fig-cap: "Simulated typical-patient concentration-time profile."
typ <- sim_typical(pars, dose = dose)
sim_plot_typical(typ, dose)
```

```{r typ-summary}
#| include: false
cmax <- round(max(typ$cp), 2)
tmax <- round(typ$time[which.max(typ$cp)], 2)
```

Predicted **Cmax ≈ `r cmax` mg/L** at **Tmax ≈ `r tmax` h**, in line with the
observed data and the NCA — the expected internal coherence of a sound model.

## Effect of body weight (illustrative)

Because the selected model has no weight covariate (it wasn't supported), this is an
**illustrative** exercise: we apply the **standard theoretical allometric exponents**
(0.75 on CL, 1.0 on V) to the typical parameters to show the *expected* direction and
size of a body-weight effect at a fixed dose. It is not an estimated effect.

```{r weight}
#| fig-cap: "Illustrative typical profiles for 55, 70, and 85 kg subjects (theory-based allometry)."
wts <- sim_weight_subgroups(pars, dose = dose)
sim_plot_weight(wts, dose)
```

```{r weight-table}
wts |>
  dplyr::group_by(WT) |>
  dplyr::summarise(Cmax = round(max(cp), 2),
                   Tmax = round(time[which.max(cp)], 2),
                   .groups = "drop") |>
  kbl(caption = "Illustrative Cmax / Tmax by body weight (fixed dose, allometric theory)",
      col.names = c("Weight (kg)", "Cmax (mg/L)", "Tmax (h)")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## Population variability

Drawing 500 virtual subjects' random effects from the estimated ω matrix shows the
**spread** of real-world exposures, not just the average. For a narrow-therapeutic-index drug like theophylline, this band is what matters — it indicates how many
patients might stray outside the therapeutic window on a given dose.

```{r population}
#| fig-cap: "Population simulation: median and 90% prediction interval (500 subjects)."
pop <- sim_population(pars, dose = dose)
if (!is.null(pop)) {
  sim_plot_population(pop, dose)
} else {
  cat("Population simulation requires an omega matrix (IIV model).")
}
```

## Key takeaways

- The qualified **IIV model** reproduces a sensible typical profile
  (Cmax ≈ `r cmax` mg/L), confirming internal consistency with the data.
- An **illustrative** allometric scaling shows the expected body-weight effect on
  exposure, clearly separated from the (unsupported) estimated covariate of Phase 5.
- The **population prediction interval** quantifies between-subject variability — the
  critical output for dosing a narrow-therapeutic-index drug safely.
- This completes the end-to-end workflow: data → NCA → structural model → IIV →
  covariate → diagnostics → simulation.

> The full analysis is published as a Quarto website under `docs/` and deployed via
> GitHub Pages.
```
