---
title: "Phase 7 — Clinical Simulation"
subtitle: "Typical patient, weight subgroups, and population variability (rxode2)"
author: "Pramod BR"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    number-sections: true
    fig-width: 8
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
here::i_am("reports/07_Simulation_Report.qmd")
source(here::here("R", "07_simulation.R"))
library(kableExtra)
```

## Objectives

With a model that passed diagnostics (Phase 6), we can now **simulate** — predict
concentration-time profiles for scenarios we never measured. This is where a
population PK model earns its keep clinically. We simulate three things with
`rxode2`: a typical patient, the effect of body weight, and the full population
spread.

```{r pars}
pars <- get_typical_params()
```

The simulations use the typical parameters from the final model
(`r pars$source`): CL ≈ `r round(pars$cl, 2)` L/h, V ≈ `r round(pars$v, 1)` L,
Ka ≈ `r round(pars$ka, 2)` /h, with weight exponents
θ_CL = `r round(pars$dWTCL, 2)`, θ_V = `r round(pars$dWTV, 2)`.

## Typical patient

Setting the random effects to zero gives the single population-average curve for a
70 kg subject after a `r SIM_DOSE` mg oral dose.

```{r typical}
#| fig-cap: "Simulated typical-patient concentration-time profile."
typ <- sim_typical(pars)
sim_plot_typical(typ)
```

```{r typ-summary}
#| include: false
cmax <- round(max(typ$cp), 2)
tmax <- round(typ$time[which.max(typ$cp)], 2)
```

Predicted **Cmax ≈ `r cmax` mg/L** at **Tmax ≈ `r tmax` h** — consistent with the
observed data and the NCA, the expected internal coherence of a good model.

## Effect of body weight

Same `r SIM_DOSE` mg dose, three weights. Because clearance and volume scale
allometrically with weight, heavier subjects show a different exposure profile —
the visual rationale behind weight-based dosing decisions.

```{r weight}
#| fig-cap: "Typical profiles for 55, 70, and 85 kg subjects at a fixed dose."
wts <- sim_weight_subgroups(pars)
sim_plot_weight(wts)
```

```{r weight-table}
wts |>
  dplyr::group_by(WT) |>
  dplyr::summarise(Cmax = round(max(cp), 2),
                   Tmax = round(time[which.max(cp)], 2),
                   .groups = "drop") |>
  kbl(caption = "Simulated Cmax / Tmax by body weight (fixed dose)",
      col.names = c("Weight (kg)", "Cmax (mg/L)", "Tmax (h)")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## Population variability

Drawing 500 virtual subjects' random effects from the estimated ω matrix shows the
**spread** of real-world exposures, not just the average. For a narrow-therapeutic-
index drug like theophylline, this band is what matters — it indicates how many
patients might stray outside the therapeutic window on a given dose.

```{r population}
#| fig-cap: "Population simulation: median and 90% prediction interval (500 subjects)."
pop <- sim_population(pars)
if (!is.null(pop)) {
  sim_plot_population(pop)
} else {
  cat("Population simulation requires an omega matrix (IIV/covariate model).")
}
```

## Key takeaways

- The qualified model reproduces a sensible **typical profile**
  (Cmax ≈ `r cmax` mg/L), confirming internal consistency with the data.
- **Body weight** measurably shifts exposure at a fixed dose — quantitative support
  for individualised dosing.
- The **population prediction interval** quantifies between-subject variability,
  the critical output for dosing a narrow-therapeutic-index drug safely.
- This completes the end-to-end workflow: data → NCA → structural model → IIV →
  covariate → diagnostics → simulation.

> The full analysis is published as a Quarto website under `docs/` and deployed via
> GitHub Pages.
```
