---
title: "Phase 9 — One- vs Two-Compartment Comparison"
subtitle: "Structural model selection: is a peripheral compartment justified?"
author: "Pramod BR"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    number-sections: true
    fig-width: 8
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
here::i_am("reports/08_Two_Compartment_Report.qmd")
source(here::here("R", "08_two_compartment.R"))
library(kableExtra)
```

## Objectives

The whole analysis assumed a **one-compartment** structure. Here we test that
assumption directly by fitting a **two-compartment** model and asking whether the
extra peripheral compartment is statistically justified. Demonstrating that a more
complex model can be fitted *and correctly accepted or rejected* is core
model-selection judgement.

## The two-compartment model

A peripheral compartment lets drug distribute into tissues and return, producing a
biphasic semilog decline. Two new parameters appear: **Q** (inter-compartmental
clearance) and **V2** (peripheral volume); the central volume becomes **Vc**.

The fitted model's structure is shown as equations in the next section.

## Fit and compare

```{r fit}
#| cache: true
dat     <- build_nonmem_data()
iiv_fit <- load_iiv_fit()
two_fit <- fit_two_cmt_model(dat)
```

```{r params}
base_param_table(two_fit) |>
  kbl(caption = "Two-compartment model estimates", digits = 3) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

The key diagnostic for *worth* is whether **Q** and **V2** are precisely estimated
(reasonable %RSE) — if their relative standard errors are very large, the data
cannot support a peripheral compartment.

## Model selection

Adding two parameters is justified (p < 0.05) only if the OFV drops by more than
**5.99** (χ², 2 df); AIC/BIC additionally penalise complexity.

```{r compare}
cmp <- compare_models(
  model_metrics(iiv_fit, "1-compartment (selected)"),
  model_metrics(two_fit, "2-compartment")
)
cmp |>
  kbl(caption = "1- vs 2-compartment model comparison", digits = 2) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r verdict}
#| include: false
dofv <- cmp$dOFV[2]; daic <- cmp$dAIC[2]
justified <- !is.na(dofv) && dofv < -5.99 && !is.na(daic) && daic < 0
```

Here ΔOFV = **`r round(dofv, 2)`** and ΔAIC = **`r round(daic, 2)`**. The
two-compartment model is therefore **`r if (justified) "supported" else "not supported"`**:
`r if (justified) "the peripheral compartment improves the fit beyond the complexity penalty." else "the extra compartment does not improve the fit enough to justify two more parameters."`

## Key takeaways

- A full **two-compartment** oral model was fitted and compared head-to-head with
  the selected one-compartment model.
- Decision by **ΔOFV / ΔAIC / BIC**: the two-compartment model is
  **`r if (justified) "preferred" else "rejected in favour of the simpler 1-compartment model"`**.
- For theophylline, the one-compartment structure is well established, so rejecting
  the extra compartment is the expected and **parsimonious** result — confirming the
  structural choice made back in Phase 3.

> Next (Phase 10): **bootstrap confidence intervals** on the final model to quantify
> parameter uncertainty empirically.
```
