---
title: "Phase 10 — Bootstrap Confidence Intervals"
subtitle: "Empirical parameter uncertainty by subject resampling"
author: "Pramod BR"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    number-sections: true
    fig-width: 8
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
here::i_am("reports/09_Bootstrap_Report.qmd")
source(here::here("R", "09_bootstrap.R"))
library(kableExtra)
```

## Objectives

The parameter CIs reported so far are **asymptotic** — derived from the covariance
matrix under large-sample, normal assumptions. With only 12 subjects those
assumptions are questionable. This phase computes **bootstrap** CIs that make no
such assumption, and compares the two.

## Method — case (subject) resampling

We repeatedly draw 12 subjects **with replacement** from the original 12, refit the
IIV model to each resampled dataset, and collect the fixed effects. The 2.5th and
97.5th percentiles of those estimates form an empirical 95% CI.

```{r run-boot}
#| cache: true
#| output: false
dat  <- build_nonmem_data()
fit  <- load_iiv_fit()
boot <- run_bootstrap(dat, nboot = 50)
```

```{r conv}
#| include: false
rate <- round(100 * boot$n_success / boot$n_attempt)
```

**`r boot$n_success` of `r boot$n_attempt`** replicates converged
(`r rate`%). A convergence rate below 100% is normal — some resamples produce
awkward datasets (e.g. the same subject drawn several times) — and the rate itself
is a useful stability check.

## Asymptotic vs bootstrap 95% CI

```{r summary}
boot_summary(boot, fit) |>
  kbl(caption = "Point estimate with bootstrap median and 95% percentile interval",
      digits = 3) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

Read this against the asymptotic CIs from Phase 4. **Agreement** between the two
reassures us the asymptotic SEs were trustworthy; a **wider or skewed** bootstrap
interval (most likely for **Ka**, the least precise parameter) reveals uncertainty
the formula understated.

## Bootstrap distributions

```{r boot-fig}
#| fig-cap: "Bootstrap sampling distributions of CL, V, and Ka (dashed = 95% percentile interval)."
boot_plot(boot)
```

## Key takeaways

- Parameter uncertainty was quantified **empirically** by subject resampling,
  free of the normality assumption behind asymptotic SEs.
- The bootstrap and asymptotic CIs can be compared directly; close agreement
  validates the reported precision, while divergence (typically for Ka) flags
  small-sample fragility.
- With n = 12 the bootstrap is itself coarse and presented as a **demonstration of
  method** — exactly how it would be framed for a small study.

> Next (Phase 11): a **covariate forest plot** summarising the weight effect on
> CL and V with confidence intervals.
```
