---
title: "Phase 11 — Covariate Forest Plot"
subtitle: "Clinical relevance of the body-weight effect on CL and V"
author: "Pramod BR"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    number-sections: true
    fig-width: 9
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
here::i_am("reports/10_Forest_Plot_Report.qmd")
source(here::here("R", "10_forest_plot.R"))
library(kableExtra)
```

## Objectives

A statistical test (Phase 5) tells us whether a covariate effect is *real*; a
**forest plot** tells us whether it is *clinically relevant*. This phase summarises
the body-weight effect on clearance and volume as parameter ratios — relative to a
typical 70 kg subject — for a light and a heavy patient, each with a 95% CI.

## How to read it

Each row shows a parameter ratio: **1.0 means no change** from the typical patient.
The shaded band (**0.8–1.25**) is the conventional "no clinically relevant effect"
zone (within ±20–25%).

- A point **and its whole CI inside the band** → the covariate doesn't move that
  parameter enough to matter clinically.
- A **CI crossing 1.0** → the effect isn't statistically distinguishable from zero.
- A tight CI **far from 1.0** → a covariate worth acting on (dose adjustment).

The ratio at weight *w* is `(w/70)^θ`, and each CI is propagated from the exponent's
95% confidence interval.

## The numbers

```{r data}
fit <- load_cov_fit()
fd  <- forest_data(fit)
fd |>
  kbl(caption = "Weight effect on CL and V — ratio vs typical 70 kg subject (95% CI)",
      digits = 3,
      col.names = c("Parameter", "Scenario", "Ratio", "Lower 95%", "Upper 95%")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## The forest plot

```{r forest}
#| fig-cap: "Forest plot of the body-weight effect on CL and V (light 55 kg, heavy 85 kg)."
forest_plot(fd)
```

## Interpretation

The **clearance** intervals comfortably **cross 1.0**, so weight is not
distinguishable from "no effect" on CL. The **volume** intervals *just* exclude 1.0
— a small effect, consistent with the marginally non-zero V exponent in Phase 5 —
but they remain **within the 0.8–1.25 band**, so even where an effect is detectable
it is **not clinically relevant**. Together with the Phase 5 likelihood-ratio test
(ΔOFV = −3.65, short of the 5.99 cut-off for the two exponents jointly), the verdict
is that body weight does **not warrant a dose adjustment** in this dataset.

## Key takeaways

- The forest plot translates a covariate effect into the question clinicians
  actually care about: *does it change exposure enough to change dosing?*
- For Theophylline weight here, the answer is **no** — the intervals sit at or within
  the no-effect band (CL spans 1.0; V shows only a small, sub-clinical effect).
- Reporting a covariate honestly as **not clinically relevant**, with a clear
  figure, is sound pharmacometric practice — and a fitting close to the portfolio.

> This completes the extended portfolio: EDA → NCA → base → IIV → covariate →
> diagnostics → simulation → 2-compartment check → bootstrap → forest plot.
```
